White Wines by Victor Jaen
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Loading libraries
library(tidyverse) #this one auomatically loads ggplot and dplyr
library(ggplot2)
library(dplyr)

## A note on syntaxis: I will be using mainly data pipes in constructing my
## plots. I know this wasnt really shown in the course but I found it easier
## to understand and work with too.

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
#setwd("C:/Users/Victor/Desktop/NAND/EDA/Final_Project") #<- Delete before Knittting
setwd("~/Documents/GitHub/Final_Project") #For Mac
df <- read.csv("wineQualityWhites.csv")
```

This data set contains several metrics about white wines and a qualitative 
variable that contains the evaluation of the whine by experts. So right from 
the start most of the analysis will be oriented in trying to identify what
variables are the most important predictor in wine quality. Besides this we will
be exploring if within this variable, some relationship exist and hypothesize
why this could be.

# Univariate Plots Section

> **Tip**: In this section, you should perform some preliminary exploration of
your dataset. Run some summaries of the data and create univariate plots to
understand the structure of the individual variables in your dataset. Don't
forget to add a comment after each plot or closely-related group of plots!
There should be multiple code chunks and text sections; the first one below is
just to help you get started.

```{r echo=FALSE, Univariate_Plots}

head(df)


```
First I print the head of the data set to have a quick glance.

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(df)

```

We can see that all variables are numerical, and I will probably use them as they are. If the need for a categorical one arises I will create it down the line. The other observation is that they have very different ranges and distributions which is expected given the variety of data we have here. All variables seem to be continuous, except for quality, which seems discrete.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=quality)) + geom_histogram()


```

I was first interested in quality. The distribution is discrete, but this is
expected since the evaluation was given from a discrete range. I will create an
additional variable which will divide the wine in bad, average, and good based
on the quality. The ranges will be from 0 to 4, 5 to 7 and then 8 to 10
respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df <-df %>% mutate(quality_group = case_when(quality <= 4 ~ "bad", 
                                      quality > 4 & quality <=7 ~ "average",
                                      quality > 7 & quality <=10 ~ "good"))

table(df$quality_group)

```

The majority of the wines are average as shown in the histogram. This will
help in plotting downstream.



```{r echo=FALSE, message=FALSE, warning=FALSE}

#To create a tidy data set we use the gather function.
df_tidy <- df %>% select(-quality_group) %>% gather(fixed.acidity,volatile.acidity, citric.acid,
                        residual.sugar,chlorides,free.sulfur.dioxide,
                        total.sulfur.dioxide,density,pH,sulphates,alcohol,
                        quality)
colnames(df_tidy) <- c("X","key","values")
df_tidy %>% ggplot(aes(x=values)) + geom_histogram(bins = 50) + facet_wrap(~key, scales= "free")

```

I wanted to have quick distribution for all variables. So using the gather
function from tidy verse, I created a tidy data set to make the plotting easier.
Most of the variables seem to be normally distributed. They had different
spreads and a little bit of skewness, but most seem pretty central. 
Residual sugar is a little skewed so I will have a look on that one.


```{r}

df %>% ggplot(aes(x=log(residual.sugar))) + geom_histogram(bins = 50)

```

Using a logarithmic scale we can see actually almost two distributions with 
a peak close to 0. We are going to have a closer look to this phenomena in
the coming sections.







# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

Is mostly of numeric variables, most of them seem to be normally distrbuted 
with different spreads.

### What is/are the main feature(s) of interest in your dataset?

If this uniformity is represented with looking a two variables simultaneously
Is there any sweet spot in this variables explained in good wine quality.

### What other features in the dataset do you think will help support your \
#investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

Yes. I created a factor variable called quality groups that divides the wine
in three main categories: bad, average and good. This will help identify 
better trends in the following plots.

### Of the features you investigated, were there any unusual distributions? \
Not really. Only the residual sugar, a little bit skewed but with a logarithmic
transformation, we saw that there seems to be to different distributions for 
that variable.

### Did you perform any operations on the data to tidy, adjust, or change the form \
### of the data? If so, why did you do this?
I created a tidy data set to leverage facet_wrap from ggplot and generate
all the distributions in one line. This tidy data set is call df_tidy.
Depending on the need I will point out if I need to use it in other plots,
however most of the analysis will be done on the main one.

# Bivariate Plots Section


```{r echo=FALSE, Bivariate_Plots}

df %>% ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide)) + geom_jitter(alpha=.5) #also confirming

```



```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=as.factor(quality),y=density)) + geom_boxplot(alpha=.5) #the higher de quality the more uniform, could be a function of qty


```

It seems that the "better" wines have lower density, given the similarity of the
7 and 8 quality. Im cautious with the 9 boxplot, since it could be
the result of less samples rather than an actual pattern. Just to confirm if 
this is the case, I will replicate the graph using the quality_group variable
created before.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=quality_group,y=density)) + geom_boxplot(alpha=.5) #the higher de quality the more uniform, could be a function of qty


```
Yes, it seems that good wines have lower density than other ones.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=total.sulfur.dioxide,y=sulphates)) + geom_point(alpha=.5)

```
Looking at other variables, I was curious if total sulfur dioxide had some
relationship with Sulphates. In this graphs there seems to be a very weak
correlation, at least weaker than I expected. (I know chemically they are
different compounds, but they usually have some relationship)


```{r}

df %>% ggplot(aes(x=fixed.acidity,y=citric.acid)) + geom_jitter(alpha=.5) 

```
Looking at the acidity variables, there seems to be some relationship
between the citric acid and the fixed acidity. Chemically Im not sure what
could be the mechanism happening behind but this seems to be expected.
Source (https://www.winespectator.com/drvinny/show/id/Difference-Between-Sulfites-Sulfates-Wine-54706)

```{r}

df %>% ggplot(aes(x=fixed.acidity,y=pH)) + geom_jitter(alpha=.5) #strongest relationship

```
Following the same principle of looking at different variables from the 
output, I wanted to confirm the relationship between pH an acidity. From
This graph we can show that there is a decent relationship. 

```{r}

cor(df$fixed.acidity,df$pH)

```
Running the correlation function we see that the value is actually rather low,
-0.42.

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
I didnt want to focus solely on the quality variable and its relationship
with other ones. Altough I recognize this may be the most important thing to do.
Having said that, I picked density to see if it showed some relationship
and it seems it does. Good wines have a lower density distribution than the 
other wines. 

### Did you observe any interesting relationships between the other features \
The acidity variables seems to have some relationship. However the sulfate
and sulphates variables does not seem to be related.

### What was the strongest relationship you found?
Density seems to have an inverse relationship with wine quality. Good wines, 
above 7, have lower densities than the average and bad wines.

# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}
df%>% ggplot(aes(x=alcohol,y=residual.sugar)) + geom_jitter(aes(color=quality),alpha=0.3)
df%>% ggplot(aes(x=alcohol,y=density)) + geom_jitter(aes(color=quality),alpha=0.3)
df%>% filter(density <= 1.02) %>% ggplot(aes(x=alcohol,y=density)) + geom_jitter(aes(color=quality),alpha=0.3)

#create another column for above 7 and below 7.
df_g<- df%>% mutate(quality_groups = ifelse(quality>=7,'good','not_good'))

df_g%>% filter(density <= 1.02) %>% ggplot(aes(x=alcohol,y=density)) + geom_jitter(aes(color=quality_groups),alpha=0.3)
df_g%>% ggplot(aes(x=citric.acid,y=alcohol)) + geom_jitter(aes(color=quality_groups),alpha=0.3)


```

```{r}

df_g%>% ggplot(aes(x=sulphates,y=chlorides)) + geom_jitter(aes(color=quality_groups),alpha=0.3)

df_g%>% ggplot(aes(x=sulphates,y=residual.sugar)) + geom_jitter(aes(color=quality_groups),alpha=0.3)



```



# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.
Exploring all variables directly with the output variable is not necessary
the best path always. Understanding relationships between variables could help
in feature selection, dropping variables that might not be interesting.



> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!