White Wines by Victor Jaen
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Loading libraries
library(tidyverse)
library(ggplot2)
library(dplyr)

## A note on syntaxis: I will be using mainly data pipes in constructing my
## plots. I know this wasnt really shown in the course but I found it easier
## to understand and work with too.

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv("wineQualityWhites.csv")

```

This data set contains several metrics about white wines and a qualitative 
variable that contains the evaluation of the whine by experts. So right from 
the start most of the analysis will be oriented in trying to identify what
variables are the most important predictor in wine quality. Besides this we will
be exploring if within this variable, some relationship exist and hypothesize
why this could be.

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}

head(df)


```
First I print the head of the data set to have a quick glance.

```{r echo=FALSE, message=FALSE, warning=FALSE}

summary(df)

```

We can see that all variables are numerical, and I will probably use them as they are. If the need for a categorical one arises I will create it down the line. The other observation is that they have very different ranges and distributions which is expected given the variety of data we have here. All variables seem to be continuous, except for quality, which seems discrete.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=quality)) + geom_histogram()


```

I was first interested in quality. The variable is discrete, but this is
expected since the evaluation was given from a discrete range. I will create an
additional variable which will divide the wine in bad, average, and good based
on the quality. The ranges will be from 0 to 4, 5 to 7 and then 8 to 10
respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df <-df %>% mutate(quality_group = case_when(quality <= 4 ~ "bad", 
                                      quality > 4 & quality <=7 ~ "average",
                                      quality > 7 & quality <=10 ~ "good"))

table(df$quality_group)

```

The majority of the wines are average as shown in the histogram. This will
help in plotting downstream.

```{r message=FALSE, warning=FALSE}

#To create a tidy dataset we use the gather function.
df_tidy <- df %>% select(-quality_group) %>% gather(key,values,fixed.acidity,volatile.acidity, citric.acid,
                        residual.sugar,chlorides,free.sulfur.dioxide,
                        total.sulfur.dioxide,density,pH,sulphates,alcohol,
                        quality)

df_tidy %>% ggplot(aes(x=values)) + geom_histogram(bins = 50) + facet_wrap(~key, scales= "free")

```

I wanted to have quick distribution of all variables. So using the gather
function from tidy verse, I created a tidy data set to make the use of
ggplot easier.Most of the variables seem to be normally distributed. 
They had different spreads and a little bit of skewness, but most seem pretty 
central. I will take a further look on a couple of variables that I think
need closer examination.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=log(residual.sugar))) + geom_histogram(bins = 50)

```

Residual sugar was very skewed to one side. Using a logarithmic scale we can see 
actually almost two distributions with peaks around 0.5 and 2. We are going to
have a closer look to this phenomena in the coming sections.


```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=chlorides)) + geom_histogram(bins=50) + scale_x_continuous(limits = c(0,0.1))

```
Examining closer the chlorides variable shows a fairly normal distribution. 
The median seems to be at 0.05.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=alcohol)) + geom_histogram(bins = 75)

```
The alcohol variable shows are sort of discrete behavior. It seems that
the variable simply is missing some specific values, they are randomly
distributed. Thinking why this could be my guess is a function of the chemical
test done to measure this variable or the instruments used.

```{r}

df %>% ggplot(aes(x=pH)) + geom_histogram(bins = 50)

```

Observing pH now that shows a couple of different peaks but still is normally
distrubted around 3.15 pH with two medians. 


# Univariate Analysis

### What is the structure of your dataset?

Is mostly of numeric variables, most of them seem to be normally distrbuted 
with different spreads.

### What is/are the main feature(s) of interest in your dataset?

If this uniformity is represented with looking a two variables simultaneously
Is there any sweet spot in this variables explained in good wine quality.

### What other features in the dataset do you think will help support your \
#investigation into your feature(s) of interest?
All the other variables. At this stage is to early to disregard any variable.

### Did you create any new variables from existing variables in the dataset?

Yes. I created a factor variable called quality groups that divides the wine
in three main categories: bad, average and good. This will help identify 
better trends in the following plots.

### Of the features you investigated, were there any unusual distributions? \
Not really. Only the residual sugar, a little bit skewed but with a logarithmic
transformation, we saw that there seems to be two different distributions for 
that variable.

### Did you perform any operations on the data to tidy, adjust, or change the form \
### of the data? If so, why did you do this?
I created a tidy data set to leverage facet_wrap from ggplot and generate
all the distributions in one line. This tidy data set is call df_tidy.
Depending on the need I will point out if I need to use it in other plots,
however most of the analysis will be done on the main one.

# Bivariate Plots Section


```{r echo=FALSE, Bivariate_Plots}

df %>% ggplot(aes(x=free.sulfur.dioxide,y=total.sulfur.dioxide)) + geom_jitter(alpha=.5) #also confirming

```



```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=as.factor(quality),y=density)) + geom_boxplot(alpha=.5) + 
  scale_y_continuous(limits = c(0.98,1.01))

```

It seems that the "better" wines have lower density, given the similarity of the
7 and 8 quality. Im cautious with the 9 boxplot, since it could be
the result of less samples rather than an actual pattern. Just to confirm if 
this is the case, I will replicate the graph using the quality_group variable
created before.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=quality_group,y=density)) + geom_boxplot(alpha=.5) + 
  scale_y_continuous(limits = c(0.98,1.01))

```
Yes, it seems that good wines have lower density than other ones.
An important thing is that this could be result of sample size, there are
less good quality wines, hence the distribution is more uniform.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=total.sulfur.dioxide,y=sulphates)) + geom_point(alpha=.5)

```
Looking at other variables, I was curious if total sulfur dioxide had some
relationship with Sulphates. In this graphs there seems to be a very weak
correlation, at least weaker than I expected. (I know chemically they are
different compounds, but they should have some relationship)


```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=fixed.acidity,y=citric.acid)) + geom_jitter(alpha=.5) 

```

Looking at the acidity variables, there seems to be some relationship
between the citric acid and the fixed acidity. Chemically Im not sure what
could be the mechanism happening behind but this seems to be expected.
Source (https://www.winespectator.com/drvinny/show/id/Difference-Between-Sulfites-Sulfates-Wine-54706)

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=chlorides,y=fixed.acidity)) + geom_jitter(alpha=.5) + 
  scale_x_continuous(limits=c(0,0.1))

```
Exploring othe variables, chlorides does not seem to have any relation with
fixed acidity.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=chlorides,y=free.sulfur.dioxide)) + geom_jitter(alpha=.5) + scale_x_continuous(limits = c(0,0.1)) + scale_y_continuous(limits=c(0,150))

```
Similar as the previous graph, there is not much going on between chlorides and
free sulfur dioxide.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=quality_group,y=chlorides)) + geom_boxplot() + scale_y_continuous(limits = c(0,0.1))

```

If chlorides are not showing any pattern with other variables I thought it 
might worth to compare it to the output variable. And indeed as density
it seems to be a pattern with good wines having lower chlorides levels.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=fixed.acidity,y=pH)) + geom_jitter(alpha=.5)

```
Continuing exploring different variables from the output, I wanted to confirm 
the relationship between pH an acidity. From This graph we can show that there 
s a decent relationship. 

```{r echo=FALSE, message=FALSE, warning=FALSE}

cor(df$fixed.acidity,df$pH)

```
Running the correlation function we see that the value is actually rather low,
-0.42. Possibly some outliers are impacting this value.

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I didnt want to focus solely on the quality variable and its relationship
with other ones. Altough I recognize this may be the most important thing to do.
Having said that, I picked density to see if it showed some relationship
and it seems it does. Good wines have a lower density distribution than the 
other wines. 

### Did you observe any interesting relationships between the other features \

The acidity variables seems to have some relationship. However the sulfate
and sulphates variables does not seem to be related. Chlorides didn't show
relationship with other variables.

### What was the strongest relationship you found?

Density seems to have an inverse relationship with wine quality. Good wines, 
above 7, have lower densities than the average and bad wines. Chlorides also
seem to show strong relationship with quality.

# Multivariate Plots Section
```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}

df %>% ggplot(aes(x=log10(residual.sugar),y=density)) + geom_smooth(aes(color=quality_group)) + scale_y_continuous(limits = c(0.985,1.005))

```
I was first curios about residual sugar and density since both showed some sort
of different relationship vs the rest of the variables. Using geom_smooth there
seems to be some sort of proportional relationship between density and the log 
of residual sugar, and it seems similar across different wine groups. The fact
that the average group goes farther is simply more variability within that group.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=fixed.acidity,y=citric.acid,color=pH)) +  geom_jitter(alpha=0.4) + scale_color_gradientn(colours = terrain.colors(10))

```
Also following up with pH, as expected there is a relationship between both
acidities and pH. The higher the acidity de lower the pH. This may seem
obvious however is always good to confirm in the dataset that the realworld
phenonomenas hold true.


```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=total.sulfur.dioxide,y=alcohol,color=quality)) + facet_wrap(~quality_group) + geom_point() + scale_color_gradientn(colours = terrain.colors(8))

```
Also following up with the sulfur impact, I was interested in seeing it compared
to alcohol. There does not seems to exist a relationship between alcohol and sulfur
dioxide. Rather some sort of limitation, higher content wine seem to have no more
than 200 of sulfur dioxide. On the quality side, high quality wines seem
be concentrated in a narrower range of sulfur, a around above 10 of alchol content
but besides does limit it could vary.

```{r echo=FALSE, message=FALSE, warning=FALSE}

df %>% ggplot(aes(x=chlorides,y=density)) + geom_point(aes(color=quality_group)) +  facet_wrap(~quality_group) + scale_x_continuous(limits = c(0,0.1)) +
   scale_y_continuous(limits = c(0.98,1.01)) 

```

Closing up density and chlorides show a relationship between them relatively
similar among quality groups.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Alcohol, density, pH and chlorides seem to be the most important ones.


### Were there any interesting or surprising interactions between features?
Actually no. It seems very spread.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.
No model was created.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}

df %>% ggplot(aes(x=density,y=alcohol,color=quality_group)) + geom_point(alpha=0.5) + facet_wrap(~quality_group) + scale_x_continuous(limits = c(0.99,1.005),breaks = seq(0.99,1.0,0.005)) + ggtitle("Alcohol vs Density grouped by quality group") + xlab("Density (g / cm^3)") + ylab("Alcohol (% volume)")

```

### Description One
Most good wines seem to have at least 10% alcohol by volume and have a narrower
band of density between 0.99 g/cm^3 and 1 g/cm^3. For the other groups, the 
variability seems much broader. In other words, good wines seem to have some 
specific limits of density and alcohol. Another reason for this graph is to
show the relationship between alcohol and density. The higher the alcohol
content the lower the density, which makes sense because of the lower density of 
alcohol itself compared to water. 



### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}

df %>% ggplot(aes(x=log10(residual.sugar),y=density,color=quality_group)) +  scale_x_continuous(limits = c(0,1.5)) + scale_y_continuous(limits = c(0.985,1.005)) + ggtitle("Residual Sugar vs Density") + ylab("Density (g / cm^3)") + xlab("Log10 Residual Sugar (g / dm^3)")  + geom_smooth()

```

### Description Two
The residual sugars bimodal distribution got my attention before. This graphs
shows how the residual sugar is related to density across all wine quality
groups. However there does not seems to be a pattern that applies only
to good quality wines when it comes to the residual sugar.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}

df %>% ggplot(aes(x=log10(chlorides),y=density)) + geom_jitter(aes(color=total.sulfur.dioxide),alpha=0.5)  + 
   scale_y_continuous(limits = c(0.98,1.01)) + ggtitle("Chlorides vs Density") + ylab("Density (g / cm^3)") + xlab("Log10 Chlorides (sodium chloride - g / dm^3)") + scale_color_gradientn(colours = terrain.colors(8)) +facet_wrap(~quality_group)

```

### Description Three
Finally chlorides, density and total sulfur signaled relationships across the
analysis. This graph shows a certain pattern between the three. The higher the
density and the chlorides content the higer the sulfur. Showing this behavior
among the three quality groups we observe that the relationship holds true. We
could very cautiously say that the better the group the less dispersion.

------

# Reflection

Exploring all variables directly with the output variable is not necessary
the best path always. Understanding relationships between variables could help
in feature selection, dropping variables that might not be interesting.
After finishing the project I also realized the importance of creating a story.
In this case was more organic than actually planned, but perhaps if beforehand
I set some sort of structure the flow of graphs and analysis will probably be 
better and also for the reader will be easier to understand.

With exploratory analysis the possibilities are endless, being able to differ
whats noise from what is actually important is the key, and is not that easy.


